\magnification=1200
\nopagenumbers
\hsize=6.0truein
\vsize=9.5truein
\font\big=cmr10 scaled \magstep1
\noindent
\centerline { NBODY2 ROUTINES}
\bigskip
\settabs\+&XXXXXXXx&xXXXXXXXXXXXXX&XXXXXXX  \cr
\+&Routine &Called by &Description  \cr
\smallskip
\hrule
\smallskip
\+&BLOCK &        &Block data initialization of local labelled COMMON. \cr
\+&BODIES &OUTPUT &Output of binaries \& single particles (options 6 \& 9). \cr
\+&CHECK  &OUTPUT &Error control and restart (options 2 \& 11). \cr
\+&CMCORR &OUTPUT &Correction of c.m. coordinates \& velocities (option 18). \cr
\+&CORE   &OUTPUT &Core radius \& density centre (option 8). \cr
\+&CPUTIM &OUTPUT \& INTGRT &Elapsed CPU time in minutes (initialized in MAIN). \cr
\+&DATA   &START  &Generation of initial conditions. \cr
\+&DEFINE  &INPUT  &Definition of input parameters, options \& counters. \cr
\+&ENERGY &OUTPUT \& SCALE &Total energy. \cr
\+&ESCAPE &OUTPUT &Removal of escaping particles (option 13). \cr
\+&FPCORR &REGINT &Derivative corrections of force polynomials. \cr
\+&FPOLY1 &START \& MERGE &Regular \& irregular force \& first derivatives. \cr
\+&FPOLY2 &START \& MERGE &Second \& third force derivatives. \cr
\+&INPUT  &START  &Main parameter input. \cr
\+&INSERT &INTGRT &Insert of particle index in sequential list. \cr
\+&INTGRT &MAIN   &Decision-making \& control of integration paths . \cr
\+&LAGR   &OUTPUT &Lagrangian radii \& half-mass radius (option 7). \cr
\+&MAIN  &        &Master control flow. \cr
\+&MERGE  &MAIN  &Merging of colliding bodies (option 12). \cr
\+&MODIFY  &MAIN  &Reading modified parameters at restart. \cr
\+&MYDUMP &MAIN \& INTGRT &COMMON save or copy for restart (options 1 \& 2). \cr
\+&NBINT  &INTGRT &Irregular integration. \cr
\+&NBLIST &START \& MERGE  &Initialization of neighbour list. \cr
\+&OUTPUT &MAIN   &Main output \& optional data save. \cr
\+&RAN2  &DATA  &Portable random number generator. \cr
\+&REGINT &INTGRT &Regular integration. \cr
\+&REMOVE &ESCAPE \& MERGE &Removal of particle from COMMON tables. \cr
\+&SCALE &START &Scaling to new units. \cr
\+&SORT2  &INTGRT \& LAGR &Sequential sorting of array. \cr
\+&START  &MAIN  &Initial setup. \cr
\+&STEPI  &INTGRT &Time-step for irregular force polynomial. \cr
\+&STEPS   &START \& MERGE &Initialization of time-steps \& differences. \cr
\+&SUBSYS  &START  &Initial subsystems for merger experiment (option 17). \cr
\+&VERIFY  &INPUT  &Validation of main input parameters. \cr
\+&XTRNL &REGINT \& FPOLY1 &Force due to Plummer potential (option 15). \cr
\+&XTRNL2 &REGINT \& FPOLY1 &Force due to logarithmic potential (option 15). \cr
\+&XVPRED &OUTPUT \& MERGE  &Prediction of coordinates and velocities. \cr
\+&ZERO  &START  &Initialization of global scalars. \cr
\smallskip
\hrule
\bye
